# ![](images/DBNcare.svg){width="177"}

# DBNcare: An R Package for Dynamic Bayesian Networks in Healthcare

## Introduction

DBNcare package has the aim of creating an open source package for the Dynamic Bayesian Network framework. The current version of the package supports multinomial Dynamic Bayesian Networks with a markov order of 1 and **no missing values**.

## Current Implemented Features

-   Definition of a Dynamic Bayesian Network with markov order of 1.

-   Interactive Visualization of G_0 and G_transition.

-   Generate a Random Multinomial Dynamic Bayesian Network.

-   Sampling.

-   Parameter Learning.

-   Structure Learning.

-   Inference (Exact and Approximate).

## Examples

### Definition of a Dynamic Bayesian Network

The following code shows how to create a Dynamic Bayesian Network model. The definition require two steps:

-   Definition of G_0

-   Definition of G_transition

Once the DBN is defined it can be visualized using the visualization methods.

```{r}
library('DynamicBayesianNetwork')
library(zeallot)
```

```{r}
my_dbn <-
  empty_DBN(dynamic_nodes = c("A", "B", "C", "D", "E","F"),
            markov_order = 1)

# Creating G_0


my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('A', 't_0'),
              to = c('D', 't_0'))

my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('A', 't_0'),
              to = c('C', 't_0'))

my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('B', 't_0'),
              to = c('C', 't_0'))

my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('B', 't_0'),
              to = c('D', 't_0'))

my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('D', 't_0'),
              to = c('E', 't_0'))
my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('E', 't_0'),
              to = c('F', 't_0'))


# Creating G_transition

my_dbn <- add_arc_DBN(DBN = my_dbn,
                      from = c('A', 't'),
                      to = c('D', 't'))
my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('A', 't'),
              to = c('C', 't'))

my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('B', 't'),
              to = c('C', 't'))

my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('B', 't'),
              to = c('D', 't'))

my_dbn <- add_arc_DBN(DBN = my_dbn,
                      from = c('D', 't'),
                      to = c('E', 't'))

my_dbn <- add_arc_DBN(DBN = my_dbn,
                      from = c('E', 't'),
                      to = c('F', 't'))



my_dbn <- add_arc_DBN(DBN = my_dbn,
                      from = c('A', 't-1'),
                      to = c('A', 't'))

my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('B', 't-1'),
              to = c('B', 't'))

my_dbn <-
  add_arc_DBN(DBN = my_dbn,
              from = c('C', 't-1'),
              to = c('C', 't'))

my_dbn <- add_arc_DBN(DBN = my_dbn,
                      from = c('D', 't-1'),
                      to = c('D', 't'))

my_dbn <- add_arc_DBN(DBN = my_dbn,
                      from = c('E', 't-1'),
                      to = c('E', 't'))

my_dbn <- add_arc_DBN(DBN = my_dbn,
                      from = c('F', 't-1'),
                      to = c('F', 't'))
```

### Visualization

The package allows the user to visualize in an interactive setting G_transition and G_0 to have better control on the defined structure.

### G_transition

```{r}
plot_g_transition(my_dbn)
```

### G_0

G_0 can be visualized and a dropdown menu will help the user to navigate the defined network. In fact when a node is selected from the menu the relationships and the node itself will be highlighted.

```{r}
plot_g0(my_dbn)
```

### Sampling

Given a Dynamic Bayesian Network, the package can generate a dataset using a sampling procedure. For the sake of simplicity a simpler DBN is defined to ease the parameters setting procedure.

```{r}
# Creating a DBN
  my_dbn <-
    empty_DBN(dynamic_nodes = c("A", "B", "C"),
              markov_order = 1)
  
  # Creating G_0
  my_dbn <-
    add_arc_DBN(DBN = my_dbn,
                from = c('A', 't_0'),
                to = c('B', 't_0'))
  my_dbn <-
    add_arc_DBN(DBN = my_dbn,
                from = c('A', 't_0'),
                to = c('C', 't_0'))
  
  my_dbn <-
    add_arc_DBN(DBN = my_dbn,
                from = c('B', 't_0'),
                to = c('C', 't_0'))
  
  # Creating G_transition
  my_dbn <- add_arc_DBN(DBN = my_dbn,
                        from = c('A', 't'),
                        to = c('B', 't'))
  my_dbn <-
    add_arc_DBN(DBN = my_dbn,
                from = c('A', 't'),
                to = c('C', 't'))
  
  my_dbn <-
    add_arc_DBN(DBN = my_dbn,
                from = c('B', 't'),
                to = c('C', 't'))
  
  my_dbn <- add_arc_DBN(DBN = my_dbn,
                        from = c('A', 't-1'),
                        to = c('A', 't'))
  
  my_dbn <-
    add_arc_DBN(DBN = my_dbn,
                from = c('B', 't-1'),
                to = c('B', 't'))
  
  my_dbn <-
    add_arc_DBN(DBN = my_dbn,
                from = c('C', 't-1'),
                to = c('C', 't'))
  
  #Defining CPT G_0
  A_lv <- c('yes', 'no')
  B_lv <- c('high', 'low')
  C_lv <- c('medium', 'high')
  A_0.prob <- array(c(0.2, 0.8),
                    dim = length(A_lv),
                    dimnames = list(A_0 = A_lv))
  B_0.prob <- array(
    c(0.25, 0.75, 0.6, 0.4),
    dim = c(length(B_lv), length(A_lv)),
    dimnames = list(B_0 = B_lv, A_0 = A_lv)
  )
  C_0.prob <- array(
    c(0.1, 0.9, 0.5, 0.5, 0.3, 0.7, 0.6, 0.4),
    dim = c(length(C_lv), length(B_lv), length(A_lv)),
    dimnames = list(C_0 = C_lv, B_0 = B_lv, A_0 = A_lv)
  )
  
  # Defining CPT G_transition
  dims_A_t <- list(A_t = A_lv)
  dims_A_t[['A_t-1']] = A_lv
  # defining dims for B
  dims_B_t <- list(B_t = B_lv, A_t = A_lv)
  dims_B_t[['B_t-1']] = B_lv
  
  # defining dims for C
  dims_C_t <- list(C_t = C_lv, A_t = A_lv, B_t = B_lv)
  dims_C_t[['C_t-1']] = C_lv
  
  #defining A_t CPT
  A_t.prob = array(c(0.8, 0.2, 0.05, 0.95),
                   dim = c(length(A_lv), length(A_lv)),
                   dimnames = dims_A_t)
  # defining B_t CPT
  B_t.prob = array(
    c(0.2, 0.8, 0.6, 0.4, 0.3, 0.7, 0.1, 0.9),
    dim = c(length(B_lv), length(A_lv), length(B_lv)),
    dimnames = dims_B_t
  )
  #defining C_t CPT
  C_t.prob = array(
    c(
      0.1,
      0.9,
      0.2,
      0.8,
      0.12,
      0.88,
      0.9,
      0.1,
      0.3,
      0.7,
      0.4,
      0.6,
      0.25,
      0.75,
      0.45,
      0.55
    ),
    dim = c(length(C_lv), length(A_lv), length(B_lv), length(C_lv)),
    dimnames = dims_C_t
  )
  
  CPTs_mydbn = list(
    A_0 = A_0.prob,
    B_0 = B_0.prob,
    C_0 = C_0.prob,
    A_t = A_t.prob,
    B_t = B_t.prob,
    C_t = C_t.prob
  )
  #defining the two dbns
  fitted_my_dbn <- DBN_parameters(my_dbn, CPTs_mydbn)
```

```{r}
plot_g_transition(my_dbn)

```

```{r}
plot_g0(my_dbn)

```

### Generating a Dataset

The dbn_sampling function allows the user to sample a dataset of 6 samples and a timeseries length of 10.

```{r}
dbn_sampling(fitted_my_dbn, 10, 6)
```

### Structure Learning

Given a dataset comprised of the Sample_id (unique index for each sample), the Time (time index) and all the variables, the package allows the user to learn the parameters of a Dynamic Bayesian Network through different structure learning methods. The package dispose of score-based, constraint-based and hybrid methods.

```{r}
#dataset generation and structure learning might require some time depending on the number of samples and the length of the time series
sampled_dataset <-dbn_sampling(fitted_DBN = fitted_my_dbn, 1e+4, 5)

#learned DBN through Hill Climbing method with log-likelihood as a score
fitted_dbn_structure_hc <-StructureLearning.Network(data = sampled_dataset, algorithm = 'hc', score = 'bic')
#learned DBN through PC stable method with mutual information as conditional independence test
fitted_dbn_structure_pc <-StructureLearning.Network(data = sampled_dataset, algorithm = 'pc.stable', score = 'mi')
```

Here a summary of the structures generated via Hill climbing and PC stable methods and the structure defined by us.

```{r}
summary_DBN(fitted_dbn_structure_hc)
```

```{r}
summary_DBN(fitted_dbn_structure_pc)
```

```{r}
summary_DBN(my_dbn)
```

### Parameter Learning

Given a dataset and the structure of a Dynamic Bayesian Network, the package allows the user to learn the parameters of the model through Maximum Likelihood Estimation (MLE).

```{r}
#here the parameters are learned given the structure previously learned through hill climbing method
fitted_dbn_with_param_learning <-DBN_parameters(DBN = fitted_dbn_structure_hc, data = sampled_dataset)
```

Here we can compare the parameters learned by the MLE and the parameters set by us.

```{r}
fitted_dbn_with_param_learning
```

```{r}
fitted_my_dbn
```

## Inference

The package does not implement native algorithm for the Dynamic Bayesian Network framework but given a fitted dbn the unrolled network can be generated, unlocking all the inference methods of the bnlearn package.

### Unroll the network

```{r}
dbn_unrolled<-get_unrolled_dbn(fitted_my_dbn, 4)
```

### Exact Inference

```{r}
library(gRain)
library(bnlearn)
```

As in bnlearn Exact inference is performed using gRain tool.

```{r}
junction <- compile(as.grain(dbn_unrolled))
querygrain(junction, nodes = "C_4")$C_4
j_c_3 <- setEvidence(junction, nodes = "C_3", states = "high") 
querygrain(j_c_3, nodes = "C_4")$C_4

```

### Approximate Inference

```{r}
#performing inference using logic sampling
cpquery(dbn_unrolled, event = (C_4 == "high"), evidence = (C_3 == "high"))
```

```{r}
#performing inference using logic sampling
cpquery(dbn_unrolled, event = (C_4 == "medium"), evidence = (C_3 == "high"))
```

```{r}
cpquery(dbn_unrolled, event = (C_4 == "high"), evidence = list(C_3 = "high"), method="lw")
```

```{r}
#performing inference using likelihood weighting
cpquery(dbn_unrolled, event = (C_4 == "medium"), evidence = list(C_3 = "high"), method="lw")
```

### Forecasting

Forecasting can be performed using the following parameters: a fitted Dynamic Bayesian Network, a list containing the observations for the timepoint of interest, the number of time points you would like to forecast.

```{r}
dbn_forecasting(fitted_my_dbn,list(Time=c(0), A=c("no"), B=c("low"), C=c("medium")), 3)

```

## Create a random Multinomial Dynamic Bayesian Network

### with a specific node ordering

Using the function random_dbn the user can create a Dynamic Bayesian Network. The random generation of the network is made up of two steps:

-   In the first one, the user can specify some characteristics of the network by using the parameter *is_same* and *prob_edge.* When is_same is TRUE G_transition will have all the relationships specified in G_0. The parameter prob_edge controls the sparsity of the network.

-   In the second step, the user can specify some properties of the multinomial distributions of the nodes. The parameter fixed_cardinality allows the user to have an homogeneus network where all nodes have the same cardinality. When fixed_cardinality is TRUE max_variables_cardinality is used to specify the cardinality of all nodes, while when fixed_cardinality is FALSE a 2\<= cardinality \<= max_variables_cardinality will be sampled for each node.

```{r}
#set seed for reproducibility
set.seed(10)
#first step: a random dbn with 5 nodes
dbn_at_random<-random_dbn(LETTERS[1:5],TRUE,0.6, 0.5)
#second step: fit all nodes distributions generating a multinomial distribution with a cardinality = 2.
dbn_at_random_fitted<-fit_random_dbn(dbn_at_random, TRUE, 2)
```

For more details on how the random Dynamic Bayesian Network is generated, please read the documentation (*?random_dbn*, *?fit_random_dbn*, *?generate_dbn_random_structure*, *?generate_dbn_nodes_distributions*).
